### 2023-08-17 latex/HP2_TranslComm.tex
adapted to new files

\chapter*{Translation \& philological commentary}
\subsection*{2.1}
\begin{translation}[hp02_001]
Now, when posture is firm, the disciplined yogi whose diet is wholesome and measured, should practise breath control according to the path taught by the teacher.
\end{translation}

@@

\begin{philcomm}[hp02_001]
\emph{dṛḍho yogī} seems unlikely as \emph{dṛḍha} usually qualifies a technique, the body, etc.
It is interesting that J10, J17 and N17 have \emph{yogamārgaṃ}. The vulgate has \emph{prāṇāyāmān}.
\end{philcomm}

\subsection*{2.2}
\begin{translation}[hp02_002]
When the wind moves, everything moves and, when it is still, everything is still. Therefore, the yogī obtains the motionless state because of restraining the breath.
\end{translation}

@@

\subsection*{2.3}
\begin{translation}[hp02_003]
As long as the breath is situated in the body, there is said to be life. The emission of this is death. Therefore, one should stop the breath.
\end{translation}

@@

\subsection*{2.4}
\begin{translation}[hp02_004]
When the channels are full of impurities, the breath does not go into the central channel. How would the state beyond the mind arise? How would perfection of the body be achieved?
\end{translation}

@@

\subsection*{2.5}
\begin{translation}[hp02_005]
When entire network of channels, which is full of impurities, becomes pure, then the yogi becomes capable of holding the breath.
\end{translation}

\begin{sources}[hp02_005]
VM 95

@@

\subsection*{2.6}
\begin{translation}[hp02_006]
Therefore, one should always practice breath [retention] with a pure mind, so that the impurities situated at the sides of Suṣumṇā wither away.
\end{translation}

@@

\subsection*{2.7}
\begin{translation}[hp02_007]
Seated in lotus pose, the yogi should breathe in the air through the moon channel, hold according to his capacity and then breathe out through the sun channel.
\end{translation}

@@

\subsection*{2.8}
\begin{translation}[hp02_008]
And having drawn the breath through the sun channel, he should gradually fill the abdomen. Having done the retention as prescribed, he should then exhale through the moon channel.
\end{translation}

@@

\subsection*{2.9}
\begin{translation}[hp02_009]
Having inhaled through the [channel] by which one exhales, one should hold it without discomfort. And then he exhales through the other [channel] gently, not forcefully.
\end{translation}

@@

\subsection*{2.10}
\begin{translation}[hp02_010]
If [the yogi] takes in the breath by the moon channel, he should then exhale the restrained [breath] through the other. Having taken in the breath through the sun channel and held it, he should then exhale through the left. By this method of sun and moon, the multitudes of channels of yamis who are meditating on both orbs become pure after three months.
\end{translation}

@@

\subsection*{2.11}
\begin{translation}[hp02_011]
One should slowly practise retentions four times, at sunrise, midday, sunset and midnight, up to [a total of] eighty.
\end{translation}

@@

\subsection*{2.12}
\begin{translation}[hp02_012]
In the lowest cessation of the breath, sweating arises, in the average, shaking and in the highest [the yogi] rises up again again in lotus pose.
\end{translation}

@@

\begin{philcomm}[hp02_012]
The manuscripts diverge greatly in the second hemistrich. All have \emph{prāṇa} (instead of \emph{deha} in the VM). J10 (and others) appear to say that the breaths rise up again and again when one is seated in the lotus pose (\emph{uttiṣṭhanty uttame prāṇā baddhe padmāsane muhuḥ}). However, this is a statement about the external signs that arise in \emph{prāṇāyāma}. V1 seems to be stating that \emph{padmāsana} rises up again and again in the highest stage of holding the breath (\emph{uttiṣṭhaty uttame prāṇarodhe padmāsanaṃ muhuḥ}).

@@

\subsection*{2.13}
\begin{translation}[hp02_013]
The yogī should practise rubbing the limbs with the sweat produced by his exertion. Because of it, firmness and dexterity of the body arise.
\end{translation}

@@

\subsection*{2.14}
\begin{translation}[hp02_014]
At the beginning of the practice, food with milk and ghee is recommended. Then, when the practice has become well established, there is no need to adopt such regulations.
\end{translation}

@@

\subsection*{2.15}
\begin{translation}[hp02_015]
Just as a lion, elephant and tiger should be tamed gradually, so the breath should be trained. Otherwise, it kills the practitioner.
\end{translation}

@@

\subsection*{2.16}
\begin{translation}[hp02_016]
The end of all diseases is caused by prāṇāyāma as prescribed. As a result of inappropriate practice any disease can arise.
\end{translation}

@@

\subsection*{2.17}
\begin{translation}[hp02_017]
Hiccups, wheezing, cough, pains in the head, ears and eyes and various [other] diseases arise as a result of the breath going awry.
\end{translation}

@@

\subsection*{2.18}
\begin{translation}[hp02_018]
One should exhale correctly, inhale correctly and hold the breath correctly. Thus, one obtains success.
\end{translation}

@@

\begin{versinnote}
\tl{yuktaṃ yuktaṃ tyajed vāyuṃ yuktaṃ yuktaṃ ca pūrayet\\%122ab ; HP2.18ab
yuktaṃ yuktaṃ ca badhnīyād evaṃ siddhim avāpnuyāt || 1\\!}
\end{versinnote}
\end{sources}

@@

\subsection*{2.19}
\begin{translation}[hp02_019]
When the channels are pure, then external signs occur. Leanness and lustre of the body are certain to arise.
\end{translation}

@@

\subsection*{2.20}
\begin{translation}[hp02_020]
Holding of the breath as long as one desires, stimulating the fire, causing the internal resonance to arise and freedom from disease occur when the channels are pure.
\end{translation}

@@

\subsection*{2.21}
\begin{translation}[hp02_021]
One who has an excess of fat and phlegm should initially practise the six therapeutic interventions. However, anyone else should not practise them when their humours are in balance.
\end{translation}

@@

\subsection*{2.22}
\begin{translation}[hp02_022]
Dhauti, basti, neti, trāṭaka, naulika and kapālabhāti. [Experts] says these are the six therapeutic interventions.
\end{translation}

@@

\subsection*{2.23}
\begin{translation}[hp02_023]
These six interventions should be kept secret. Purifiers of the body, which bestow various good qualities, they are worshipped by the best yogis.
\end{translation}

@@

\subsection*{2.24}
\begin{translation}[hp02_024]
One should swallow a cloth that is four finger-breadths wide and has been soaked. Then one should draw it out. This has been taught as the action of dhauti.
\end{translation}

@@

\subsection*{2.25}
\begin{translation}[hp02_025]
Coughing, wheezing, splenitis and skin diseases, as well as the twenty phlegmatic diseases, undoubtedly disappear through the power of the dhauti technique.
\end{translation}

@@

\subsection*{2.26}
\begin{translation}[hp02_026]
In water up to the navel, one adopts a squatting pose (\emph{utkaṭa}) with a reed fixed in the anus. One should clench the base. This cleansing is bastikarma.
\end{translation}

@@

\subsection*{2.27}
\begin{translation}[hp02_027]
By the power of the basti technique, swelling, splenitis, stomach disorders and all diseases arising from wind, bile and phlegm are removed.
\end{translation}

@@

\subsection*{2.28}
\begin{translation}[hp02_028]
When practised repeatedly the jalabasti technique bestows purity of the bodily constituents, senses and mind, radiance, [and] stimulation of the digestive fire, and removes [excessive] accumulation of any humour.
\end{translation}

@@

\subsection*{2.29}
\begin{translation}[hp02_029]
[The technique] which, after the apāna wind has been raised into the tube of the throat (i.e., oesophagus), ejects the contents of the stomach from the windpipe (which has been brought under control by cumulative practice) is called the elephant’s action by experts of Haṭha.
\end{translation}

@@

\subsection*{2.30}
\begin{translation}[hp02_030]
[The yogi] should insert a thread which is handspan [long] and very smooth into the nasal passage. He takes it out through the mouth. This is called neti by the Siddhas.
\end{translation}

@@

\subsection*{2.31}
\begin{translation}[hp02_031]
And neti purifies the skull, bestows devine sight and quickly cures a multitude of diseases that arise above the clavicles.
\end{translation}

@@

\subsection*{2.32}
\begin{translation}[hp02_032]
The focused [yogi] should look at a small focal point with unmoving gaze until tears fall. This is taught as trāṭaka by teachers.
\end{translation}

@@

\subsection*{2.33}
\begin{translation}[hp02_033]
It is the destroyer of eye diseases and the door [blocking] sloth and so forth. Trāṭaka should be carefully concealed just as chest of gold.
\end{translation}

@@

\subsection*{2.34}
\begin{translation}[hp02_034]
With the shoulders lowered, one should rotate the stomach to the left and right With speed of a rapid whirlpool. This is called Nauli by people from Gauḍa.
\end{translation}

@@

\subsection*{2.35}
\begin{translation}[hp02_035]
Naulī brings about the stimulation of sluggish bodily fire, digestion and the like, always brings bliss, and makes all humoural disorders and diseases wither away. Naulī is the best of all Haṭha practices.
\end{translation}

@@

\subsection*{2.36}
\begin{translation}[hp02_036]
Like the bellows of a blacksmith, the inhalation and exhalation are fast. It is called kapālabhāti. It dries up all phlegmatic disorders.
\end{translation}

@@

\subsection*{2.37}
\begin{translation}[hp02_037]
One whose excess weight, phlegm, fat, impurity and the like have been removed by the ṣaṭkarma, should then do prāṇāyāma. It is accomplished without exertion.
\end{translation}

@@

\subsection*{2.38}
\begin{translation}[hp02_038]
Some teachers say that all the impurities are dried up by breath-controls alone and do not recommend any other practice.
\end{translation}

@@

\begin{versinnote}
\tl{prāṇāyāmair eva \textup{<2 syllables missing>} praśuṣyanti malā iti |\\+}
\tl{keṣāṃ cid ācāryāṇām anya karma na saṃmatam iti ||\\!}
\end{versinnote}
\end{testimonia}

@@

\subsection*{2.39}
\begin{translation}[hp02_039]
\end{translation}

@@

\subsection*{2.40}
\begin{translation}[hp02_040]
So long as the breath is bound in the body, the mind without support, and one looks at the middle of the brow, where is the fear of death?
\end{translation}

\begin{sources}[hp02_040]
VM

@@

\subsection*{2.41}
\begin{translation}[hp02_041]
When the network of channels has been purified by breath-controls as prescribed, the breath pierces and easily enters the mouth of suṣumṇā.
\end{translation}

@@

\subsection*{2.42}
\begin{translation}[hp02_042]
When the breath moves in the central channel, stillness of the mind arises. The mind’s becoming very still is the ‘mind beyond mind’ (\emph{manonmanī}) state.
\end{translation}

@@

\subsection*{2.43}
\begin{translation}[hp02_043]
In order to achieve that, those who know [their] methods perform various retentions. As a result of the practice of the various different retentions, [the yogi] obtains various different results.
\end{translation}

@@

\subsection*{2.44}
\begin{translation}[hp02_044]
Sūryabhedana, ujjāyī, sītkāra, [etc.] and kevala are the eight kumbhakas.
\end{translation}

@@

\subsection*{2.45}
\begin{translation}[hp02_045]
At the end of the inhalation, the lock called Jālandhara should be done. However, at the end of the retention and beginning of the exhalation, uḍḍiyāna should be done.
\end{translation}

@@

\subsection*{2.46}
\begin{translation}[hp02_046]
When the contraction of the throat has been applied, the breath goes into the central channel by quickly contracting below [the abdomen] and stretching back the middle [of the body].
\end{translation}

\begin{sources}[hp02_046]
\end{sources}

@@

\subsection*{2.47}
\begin{translation}[hp02_047]
Having raised up \emph{apānavāyu}, one should move \emph{prāṇa} down from the throat. The yogi is freed from old age and becomes sixteen years old.
\end{translation}

@@

\subsection*{2.48}
\begin{translation}[hp02_048]
Now, piercing the sun---
Having adopted vajrāsana on a comfortable mat, the yogi should gradually draw in the external air through the right nostril,
\end{translation}

@@

\subsection*{2.49}
\begin{translation}[hp02_049]
[The yogi] does \emph{kumbhaka} until the cessation [of the breath] as far as the tips of the hair and nails. The wise yogi should then exhale the breath slowly through the left nostril.
\end{translation}

@@

\subsection*{2.50}
\begin{translation}[hp02_050]
This purifies the skull, cures [imbalances] of vātadoṣa [and] gets rid of diseases caused by worms [so] should be done repeatedly. It is called the piercing of the sun.
\end{translation}

\begin{sources}[hp02_050]
Gorakṣaśataka 35a-36b

@@

\subsection*{2.51}
\begin{translation}[hp02_051]
Having closed the mouth, one should gradually draw in the breath through the nostrils so that it comes into contact with the throat as far as the chest and resonates.\end{translation}

\begin{sources}[hp02_051]
Gorakṣaśataka

@@

\subsection*{2.52}
\begin{translation}[hp02_052]
As before, he should hold the breath and then exhale through the \emph{iḍā} channel. It cures diseases caused by phlegm in the throat and increases the body’s fire.
\end{translation}

\begin{sources}[hp02_052]
GŚ

@@

\subsection*{2.53}
\begin{translation}[hp02_053]
It cures diseases in the bodily constituents inside the network of nāḍīs,
The retention called ujjāyī should be done by one moving or stationary.
\end{translation}

@@

\subsection*{2.54}
\begin{translation}[hp02_054]
Now śītkāra---
He should constantly produce a \emph{śīt} sound in the mouth and a flaring of the nostrils. By practising thus, one becomes a second god of love.
\end{translation}

@@

\subsection*{2.55}
\begin{translation}[hp02_055]
Having become one among the circle of yoginis, he can bring about creation and destruction. Neither hunger nor thirst [nor] sleep nor indolence arise.
\end{translation}

@@

\subsection*{2.56}
\begin{translation}[hp02_056]
His body is as he wishes, and he is free from all afflictions. By this method, he truly becomes the lord of yogis in the region of the earth.
\end{translation}

@@

\subsection*{2.56*1}
\begin{translation}[hp02_056_1]
One who always takes in the breath through the aperture at the roots of the tongue, undoubtedly becomes a receptacle of all \emph{siddhi}s.
\end{translation}

@@

\subsection*{2.57}
\begin{translation}[hp02_057]
It has also been taught:
He who regularly takes in the breath through the root of the tongue and palate, has all his diseases cured in half a year.
\end{translation}

@@

\emph{rasanām unmukhīkṛtya sītkāraṃ kurvatā marut} |\\
\emph{pīyante kumbhake yasmin nāsikābhyāṃ virecanam} || [pīyate?]

@@

\emph{rasanāṃ tālumūle yaḥ sthāpayitvā vipaścitaḥ} |\\
\emph{pibet prāṇānilaṃ tasya rogāṇāṃ saṃkṣayo bhavet} ||

@@

\subsection*{2.58}
\begin{translation}[hp02_058]
Having drawn in the air through the tongue, the method of retention [is done] as before. The wise yogi gradually exhales through the nostrils.
\end{translation}

@@

\subsection*{2.59}
\begin{translation}[hp02_059]
Diseases such as swelling, enlargement of the spleen and the like, and fever, [excess] bile, hunger and thirst; this retention called śītalī cures them.
\end{translation}

@@

\subsection*{2.60}
\begin{translation}[hp02_060]
Now, bhastrikā---
If [the yogi] places the soles of both feet on the thighs, the lotus pose, which destroys all bad deeds, duly arises.
\end{translation}

@@

\subsection*{2.61}
\begin{translation}[hp02_061]
Having adopted lotus pose correctly, the wise yogi whose neck and abdomen are straight should close the mouth and exhale the breath through the nose effortfully.
\end{translation}

@@

\subsection*{2.62}
\begin{translation}[hp02_062]
In such a way that [the breath] comes into contact with the chest and throat, and there is then a sound in the skull, he should quickly inhale a small amount of the breath as far as the heart lotus.
\end{translation}

@@

\subsection*{2.63}
\begin{translation}[hp02_063]
Then, the yogi should exhale and inhale again and again. In the very same way as blacksmiths’ bellows are operated forcefully, [...]
\end{translation}

@@

\subsection*{2.64}
\begin{translation}[hp02_064]
[...] the wise [yogi] should move the breath in the body. When fatigue arises in the body, he should inhale by way of the sun
\end{translation}

@@

\begin{philcomm}[hp02_064]
Most of the manuscripts support \emph{dhiyā} but the manuscripts of the source text and the testimonia support \emph{sudhīḥ}. Since the subject of the simile is \emph{bhastrā}, one would expect the subject of \emph{cālayet}, which must be different, to be stated (as is the case with \emph{sudhīḥ}). Also, one would expect the instrumental of \emph{dhī}to be qualified by some adjective, such as in the case of \emph{sattvāsthayā dhiyā} (Gorakṣaśataka 74b) and \emph{sāttvikayā dhiyā} (Haṭhapradīpikā 2.6b).
\end{philcomm}

\subsection*{2.65}
\begin{translation}[hp02_065]
in such a way that the abdomen is filled quickly by the breath and  hold the nose firmly without using the middle and index fingers.
\end{translation}

@@

\subsection*{2.78}
\begin{translation}[hp02_078]
At the end of exhaling the retained breath, one should make the mind supportless. One reaches the state of Rājayoga by practising thus.
\end{translation}

>>>>

\chapter*{Translation \& philological commentary}
%%%%%%%%%%
\subsection*{2.1}
\begin{translation}[hp02_001]
Now, when posture is firm, the disciplined yogi whose diet is wholesome and measured, should practise breath control according to the path taught by the teacher.
\end{translation}

@@

\begin{philcomm}[hp02_001]
\emph{dṛḍho yogī} seems unlikely as \emph{dṛḍha} usually qualifies a technique, the body, etc.
The vulgate has \emph{prāṇāyāmān}. Other witnesses have the plural \emph{prāṇāyāmān} and this variation between singular and plural recurs through this chapter without a clear split among different branches of the stemma, so we have preferred what seem to us to be the better readings on semantic grounds.
It is interesting that J10, J17 and N17 have \emph{yogamārgaṃ} but it is unlikely that this was the original reading.
\end{philcomm}

%%%%%%%%%%
\subsection*{2.2}
\begin{translation}[hp02_002]
When the wind moves, everything moves and, when it is still, everything is still. Therefore, the yogī obtains the motionless state because of restraining the breath.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.3}
\begin{translation}[hp02_003]
As long as the breath is situated in the body, there is said to be life. The emission of this is death. Therefore, one should stop the breath.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.4}
\begin{translation}[hp02_004]
When the channels are full of impurities, the breath does not go into the central channel. How would the state beyond the mind arise? How would perfection of the body be achieved?
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.5}
\begin{translation}[hp02_005]
When entire network of channels, which is full of impurities, becomes pure, then the yogi becomes capable of holding the breath.
\end{translation}

\begin{sources}[hp02_005]
VM 76

@@

%%%%%%%%%%
\subsection*{2.6}
\begin{translation}[hp02_006]
Therefore, one should always practice breath [retention] with a pure mind, so that the impurities situated at the sides of Suṣumṇā wither away.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.7}
\begin{translation}[hp02_007]
Seated in lotus pose, the yogi should breathe in the air through the moon channel, hold according to his capacity and then breathe out through the sun channel.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.8}
\begin{translation}[hp02_008]
And having drawn the breath through the sun channel, he should gradually fill the abdomen. Having done the retention as prescribed, he should then exhale through the moon channel.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.9}
\begin{translation}[hp02_009]
Having inhaled through the [channel] by which one exhales, one should hold it without discomfort. And then he exhales through the other [channel] gently, not forcefully.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.10}
\begin{translation}[hp02_010]
If [the yogi] takes in the breath by the moon channel, he should then exhale the restrained [breath] through the other. Having taken in the breath through the sun channel and held it, he should then exhale through the left. By this method of sun and moon, the multitudes of channels of yamis who are meditating on both orbs become pure after three months.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.11}
\begin{translation}[hp02_011]
One should slowly practise retentions four times, at sunrise, midday, sunset and midnight, up to [a total of] eighty.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.12}
\begin{translation}[hp02_012]
In the lowest cessation of the breath, sweating arises, in the average, shaking and in the highest [the yogi] rises up again again in lotus pose.
\end{translation}

@@

\begin{philcomm}[hp02_012]
The manuscripts diverge greatly in the second hemistich. All have \emph{prāṇa} (instead of \emph{deha} in the VM). J10 (and others) appear to say that the breaths rise up again and again when one is seated in the lotus pose (\emph{uttiṣṭhanty uttame prāṇā baddhe padmāsane muhuḥ}). However, this is a statement about the external signs that arise in \emph{prāṇāyāma}. V1 seems to be stating that \emph{padmāsana} rises up again and again in the highest stage of holding the breath (\emph{uttiṣṭhaty uttame prāṇarodhe padmāsanaṃ muhuḥ}).

@@

%%%%%%%%%%
\subsection*{2.13}
\begin{translation}[hp02_013]
The yogī should practise rubbing the limbs with the sweat produced by his exertion. Because of it, firmness and dexterity of the body arise.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.14}
\begin{translation}[hp02_014]
At the beginning of the practice, food with milk and ghee is recommended. Then, when the practice has become well established, there is no need to adopt such regulations.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.15}
\begin{translation}[hp02_015]
Just as a lion, elephant and tiger should be tamed gradually, so the breath should be trained. Otherwise, it kills the practitioner.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.16}
\begin{translation}[hp02_016]
The end of all diseases is caused by prāṇāyāma as prescribed. As a result of inappropriate practice any disease can arise.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.17}
\begin{translation}[hp02_017]
Hiccups, wheezing, cough, pains in the head, ears and eyes and various [other] diseases arise as a result of the breath going awry.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.18}
\begin{translation}[hp02_018]
One should exhale correctly, inhale correctly and hold the breath correctly. Thus, one obtains success.
\end{translation}

@@

\begin{versinnote}
\tl{yuktaṃ yuktaṃ tyajed vāyuṃ yuktaṃ yuktaṃ ca pūrayet\\%122ab ; HP2.18ab
yuktaṃ yuktaṃ ca badhnīyād evaṃ siddhim avāpnuyāt ||\\!}
\end{versinnote}
\end{sources}

@@

%%%%%%%%%%
\subsection*{2.19}
\begin{translation}[hp02_019]
When the channels are pure, then external signs occur. Leanness and lustre of the body are certain to arise.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.20}
\begin{translation}[hp02_020]
Holding of the breath as long as one desires, stimulating the fire, causing the internal resonance to arise and freedom from disease occur when the channels are pure.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.21}
\begin{translation}[hp02_021]
One who has an excess of fat and phlegm should initially practise the six therapeutic interventions. However, anyone else should not practise them when their humours are in balance.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.22}
\begin{translation}[hp02_022]
Dhauti, basti, neti, trāṭaka, naulika and kapālabhāti. [Experts] says these are the six therapeutic interventions.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.23}
\begin{translation}[hp02_023]
These six interventions should be kept secret. Purifiers of the body, which bestow various good qualities, they are worshipped by the best yogis.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.24}
\begin{translation}[hp02_024]
One should swallow a cloth that is four finger-breadths wide and has been soaked. Then one should draw it out. This has been taught as the action of dhauti.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.25}
\begin{translation}[hp02_025]
Coughing, wheezing, splenitis and skin diseases, as well as the twenty phlegmatic diseases, undoubtedly disappear through the power of the dhauti technique.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.26}
\begin{translation}[hp02_026]
In water up to the navel, one adopts a squatting pose (\emph{utkaṭa}) with a reed fixed in the anus. One should clench the base. This cleansing is bastikarma.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.27}
\begin{translation}[hp02_027]
By the power of the basti technique, swelling, splenitis, stomach disorders and all diseases arising from wind, bile and phlegm are removed.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.28}
\begin{translation}[hp02_028]
When practised repeatedly the jalabasti technique bestows purity of the bodily constituents, senses and mind, radiance, [and] stimulation of the digestive fire, and removes [excessive] accumulation of any humour.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.29}
\begin{translation}[hp02_029]
[The technique] which, after the apāna wind has been raised into the tube of the throat (i.e., oesophagus), ejects the contents of the stomach from the windpipe (which has been brought under control by cumulative practice) is called the elephant’s action by experts of Haṭha.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.30}
\begin{translation}[hp02_030]
[The yogi] should insert a thread which is handspan [long] and very smooth into the nasal passage. He takes it out through the mouth. This is called neti by the Siddhas.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.31}
\begin{translation}[hp02_031]
And neti purifies the skull, bestows divine sight and quickly cures a multitude of diseases that arise above the clavicles.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.32}
\begin{translation}[hp02_032]
The focused [yogi] should look at a small focal point with unmoving gaze until tears fall. This is taught as trāṭaka by teachers.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.33}
\begin{translation}[hp02_033]
It is the destroyer of eye diseases and the door [blocking] sloth and so forth. Trāṭaka should be carefully concealed just as chest of gold.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.34}
\begin{translation}[hp02_034]
With the shoulders lowered, one should rotate the stomach to the left and right With speed of a rapid whirlpool. This is called Nauli by people from Gauḍa.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.35}
\begin{translation}[hp02_035]
Naulī brings about the stimulation of sluggish bodily fire, digestion and the like, always brings bliss, and makes all humoural disorders and diseases wither away. Naulī is the best of all Haṭha practices.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.36}
\begin{translation}[hp02_036]
Like the bellows of a blacksmith, the inhalation and exhalation are fast. It is called kapālabhāti. It dries up all phlegmatic disorders.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.37}
\begin{translation}[hp02_037]
One whose excess weight, phlegm, fat, impurity and the like have been removed by the ṣaṭkarma, should then do prāṇāyāma. It is accomplished without exertion.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.38}
\begin{translation}[hp02_038]
Some teachers say that all the impurities are dried up by breath-controls alone and do not recommend any other practice.
\end{translation}

@@

\begin{versinnote}
\tl{prāṇāyāmair eva {\footnotesize <2 syllables missing>} praśuṣyanti malā iti |\\+}
\tl{keṣāṃ cid ācāryāṇām anya karma na saṃmatam iti ||\\!}
\end{versinnote}
\end{testimonia}

@@

%%%%%%%%%%
\subsection*{2.39}
\begin{translation}[hp02_039]
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.40}
\begin{translation}[hp02_040]
So long as the breath is bound in the body, the mind without support, and one looks at the middle of the brow, where is the fear of death?
\end{translation}

\begin{sources}[hp02_040]
VM 73

@@

%%%%%%%%%%
\subsection*{2.41}
\begin{translation}[hp02_041]
When the network of channels has been purified by breath-controls as prescribed, the breath pierces and easily enters the mouth of suṣumṇā.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.42}
\begin{translation}[hp02_042]
When the breath moves in the central channel, stillness of the mind arises. The mind’s becoming very still is the ‘mind beyond mind’ (\emph{manonmanī}) state.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.43}
\begin{translation}[hp02_043]
In order to achieve that, those who know [their] methods perform various retentions. As a result of the practice of the various different retentions, [the yogi] obtains various different results.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.44}
\begin{translation}[hp02_044]
Sūryabhedana, ujjāyī, sītkāra, [etc.] and kevala are the eight kumbhakas.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.45}
\begin{translation}[hp02_045]
At the end of the inhalation, the lock called Jālandhara should be done. However, at the end of the retention and beginning of the exhalation, uḍḍiyāna should be done.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.46}
\begin{translation}[hp02_046]
When the contraction of the throat has been applied, the breath goes into the central channel by quickly contracting below [the abdomen] and stretching back the middle [of the body].
\end{translation}

\begin{sources}[hp02_046]
Yogabīja ?
\end{sources}

@@

%%%%%%%%%%
\subsection*{2.47}
\begin{translation}[hp02_047]
Having raised up \emph{apānavāyu}, one should move \emph{prāṇa} down from the throat. The yogi is freed from old age and becomes sixteen years old.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.48}
\begin{translation}[hp02_048]
Now, piercing the sun---
Having adopted vajrāsana on a comfortable mat, the yogi should gradually draw in the external air through the right nostril,
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.49}
\begin{translation}[hp02_049]
[The yogi] does \emph{kumbhaka} until the cessation [of the breath] as far as the tips of the hair and nails. The wise yogi should then exhale the breath slowly through the left nostril.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.50}
\begin{translation}[hp02_050]
This purifies the skull, cures [imbalances] of vātadoṣa [and] gets rid of diseases caused by worms [so] should be done repeatedly. It is called the piercing of the sun.
\end{translation}

\begin{sources}[hp02_050]
Gorakṣaśataka 35ab, 36ab

@@

%%%%%%%%%%
\subsection*{2.51}
\begin{translation}[hp02_051]
Having closed the mouth, one should gradually draw in the breath through the nostrils so that it comes into contact with the throat as far as the chest and resonates.\end{translation}

\begin{sources}[hp02_051]
Gorakṣaśataka 36c–37b

@@

%%%%%%%%%%
\subsection*{2.52}
\begin{translation}[hp02_052]
As before, he should hold the breath and then exhale through the \emph{iḍā} channel. It cures diseases caused by phlegm in the throat and increases the body’s fire.
\end{translation}

\begin{sources}[hp02_052]
GŚ 37c–38b

@@

%%%%%%%%%%
\subsection*{2.53}
\begin{translation}[hp02_053]
It cures diseases in the bodily constituents inside the network of nāḍīs,
The retention called ujjāyī should be done by one moving or stationary.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.54}
\begin{translation}[hp02_054]
Now śītkāra---
He should constantly produce a \emph{śīt} sound in the mouth and a flaring of the nostrils. By practising thus, one becomes a second god of love.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.55}
\begin{translation}[hp02_055]
Having become one among the circle of yoginis, he can bring about creation and destruction. Neither hunger nor thirst [nor] sleep nor indolence arise.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.56}
\begin{translation}[hp02_056]
His body is as he wishes, and he is free from all afflictions. By this method, he truly becomes the lord of yogis in the region of the earth.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.56*1}
\begin{translation}[hp02_056_1]
One who always takes in the breath through the aperture at the roots of the tongue, undoubtedly becomes a receptacle of all \emph{siddhi}s.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.57}
\begin{translation}[hp02_057]
It has also been taught:
He who regularly takes in the breath through the root of the tongue and palate, has all his diseases cured in half a year.
\end{translation}

@@

\begin{versinnote}
\tl{rasanām unmukhīkṛtya sītkāraṃ kurvatā marut |\\+}
\tl{pīyante kumbhake yasmin nāsikābhyāṃ virecanam || [pīyate?]\\!}
\end{versinnote}

@@

\begin{versinnote}
\tl{rasanāṃ tālumūle yaḥ sthāpayitvā vipaścitaḥ |\\+}
\tl{pibet prāṇānilaṃ tasya rogāṇāṃ saṃkṣayo bhavet ||\\!}
\end{versinnote}

@@

%%%%%%%%%%
\subsection*{2.58}
\begin{translation}[hp02_058]
Having drawn in the air through the tongue, the method of retention [is done] as before. The wise yogi gradually exhales through the nostrils.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.59}
\begin{translation}[hp02_059]
Diseases such as swelling, enlargement of the spleen and the like, and fever, [excess] bile, hunger and thirst; this retention called śītalī cures them.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.60}
\begin{translation}[hp02_060]
Now, bhastrikā---
If [the yogi] places the soles of both feet on the thighs, the lotus pose, which destroys all bad deeds, duly arises.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.61}
\begin{translation}[hp02_061]
Having adopted lotus pose correctly, the wise yogi whose neck and abdomen are straight should close the mouth and exhale the breath through the nose effortfully.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.62}
\begin{translation}[hp02_062]
In such a way that [the breath] comes into contact with the chest and throat, and there is then a sound in the skull, he should quickly inhale a small amount of the breath as far as the heart lotus.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.63}
\begin{translation}[hp02_063]
Then, the yogi should exhale and inhale again and again. In the very same way as blacksmiths’ bellows are operated forcefully, [...]
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.64}
\begin{translation}[hp02_064]
[...] the wise [yogi] should move the breath in the body. When fatigue arises in the body, he should inhale by way of the sun
\end{translation}

@@

\begin{philcomm}[hp02_064]
Most of the manuscripts support \emph{dhiyā} but the manuscripts of the source text and the testimonia support \emph{sudhīḥ}. Since the subject of the simile is \emph{bhastrā}, one would expect the subject of \emph{cālayet}, which must be different, to be stated (as is the case with \emph{sudhīḥ}). Also, one would expect the instrumental of \emph{dhī} to be qualified by some adjective, such as in the case of \emph{sattvāsthayā dhiyā} (Gorakṣaśataka 74b) and \emph{sāttvikayā dhiyā} (Haṭhapradīpikā 2.6b).
\end{philcomm}

%%%%%%%%%%
\subsection*{2.65}
\begin{translation}[hp02_065]
in such a way that the abdomen is filled quickly by the breath and  hold the nose firmly without using the middle and index fingers.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.78}
\begin{translation}[hp02_078]
At the end of exhaling the retained breath, one should make the mind supportless. One reaches the state of Rājayoga by practising thus.
\end{translation}
